A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-20 paper.tex
Remove "of course"
It's redundant

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective. This ability can be useful in practice and could be
% Do we need to spell this out? Gregor: I think so - see below
required if, for example, we have a deeply sequenced pedigree as in \cite{RosFreixedes2022}.
% e.g. Honeybees? Be great to cite an example!
% Note by Yan - if we actually have the sequence, then this turns into
% a sample node, so would be normally included anyway. Perhaps, therefore
% we should talk about these "pass though" nodes in the context
% of internal samples.
% Gregor: There are no good deeply sequenced pedigrees in honeybees, but sex
% chromosome behaves like haplodiploid system, also any deep pedigree works here, so I cited our pig
% work. Also, I think its important to not think of all genomes as current samples - we have many
% generations of animals with genome or SNP array data and we know pedigree so we can't assume they
% are all current samples - I believe you made this clear distinction also in the tsdate paper with
% ancient human samples

@@

\section*{Ancestry resolution}\label{Ancestry_resolution}
% This is a rough first pass. Will clarify terminology later.
% Need to introduce Hudson's algorithm as concerned with simulating the process of
% the coalescent with recombination as efficiently as possible, and properly merge in
% with this section.

The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (``marginal'') trees.

We can define a sample-resolved gARG as a tuple $(S, E)$
where $S \subset \mathbb{N}$ is a set of sample nodes and
$E$ is a set of edges. Each element of $E$
is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$
over which genome $c$ inherits from $p$, and there is a path from
$c$ to some non-empty subset of $S$ through the local trees.
% Should probably define how we build the local trees somewhere.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(A) Edge annotations that reflect immediate inheritance, i.e. transmission of
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(B) Sample-resolved edge annotations: the Hudson simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{w}.
In this case, using line widths to reflect the genomic span of each edge can
help to visualise the relative importance of different transmission pathways.
}
\end{figure}

@@

Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
is given by [Rough notes]

\begin{itemize}
\item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
\item The resolved interval is then the intersection of $A_u$ and the
original inheritance interval $I_v$ on the edge.
\item It would be good of if differentiate between the inheritance intervals
 from the gARG and the ancestry intervals of the resolved gARG.
\end{itemize}

This process is illustrated in Fig~\ref{fig-ancestry-resolution}. It is intimately linked to
Hudson's algorithm which operates by
tracking segments of ancestral material carried by ancestors as
we go backwards in time.

% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.
Each lineage consists of a list of
disjoint ancestry segments $(\ell, r, a)$, where
$[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
tracking the number of samples to which the lineage is ancestral over that interval.
(We also usually track the tree node associated with each segment, but
that is not important for our purposes here so we omit it.)
If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
of the form $\{(0, m, 1)\}$. The process then works backwards in time from
the present day as a series of random common ancestor or recombination events.
Recombination events occur at a rate determined by the amount of ancestral material and
the way in which it is distributed along a chromosome.
Let $L$ be the set of ancestors at a given time $t$. Recombination events
happen at rate $\rho \nu / (m - 1)$ where
\[
\nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
    - \min_{(\ell, r, a) \in x}\ell - 1 \right)
\]
is the number of available `links' that may be broken. Thus, the rate of
recombination is determined by the left- and right-most extent of the
ancestral material carried by each lineage. At a recombination
event we choose one of these links uniformly and break it. Given a lineage
$x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
$x_1$ and $x_2$ such that FILL IN DETAILS

When $k = |L|$ lineages are present, common ancestor events
occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
are chosen uniformly at random and their ancestry segments merged.
If we have overlapping intervals of ancestry from the two lineages,
say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
\emph{coalescence} occurs and ancestor represented by the current event
will be present as a node (at least) in the marginal trees covering
the interval $[\ell, r)$. The result of this coalescence is a segment
$(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
we have found the most recent common ancestor of all samples in
the interval $[\ell, r)$ and so we do not need to simulate its history any further.
Nonoverlapping intervals of ancestry from the two lineages are included
in the resulting lineage without changes. Eventually, as the process continues,
we find resultant lineages in which all segments have fully coalescenced,
and so the number of extant lineages gradually dwindles down to zero.

@@

The state-space of the Griffiths process is much simpler than Hudson's algorithm,
which greatly facilitates mathematical reasoning. This simplicity comes at a
substantial cost, however, if we wish to use it as a practical means of
simulating recombinant ancestry. The big ARG is vast, and any realisation
for even moderate levels of recombination is far too large to be of practical
use. The number of events in the big ARG all the way back to the GMRCA
is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
of events required to simulate the little ARG is
$O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
This disparity in the number of events in the two formulations is
because the majority of the events that occur in the big ARG do
not affect the genetic ancestry of the sample in any way. Recombination
events that occur outside of ancestral material do not have any bearing
on the ancestry of the sample, and so the structure is hugely redundant.
As~\cite{wiuf1999recombination} note,
``an `ancestral' sequence in the birth and death process
need not have any genetic material in common with a
sequence descended from it.''
% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

@@

Under the coalescent with recombination (or SMC) each local
tree along the genome is transformed into the next by
a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the "subtree transfer" operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

>>>>

The focus on genomes rather than events has some interesting
and important ramifications. Firstly, a gARG may contain nodes that
do not correspond to any particular event in the ancestry of the samples.
For example, node \textsf{h} is the direct ancestor of \textsf{d} in
Fig~\ref{fig-arg-in-pedigree}, and is not the product of either a
coalescence or a recombination. Such nodes would usually be removed
from the graph so that \textsf{d} descends directly from \textsf{k} (see
the ``\nameref{ARG_simplification}'' section) but there is no necessity for this from
a representational perspective. This ability can be useful in practice and could be
% Do we need to spell this out? Gregor: I think so - see below
required if, for example, we have a deeply sequenced pedigree as in \cite{RosFreixedes2022}.
% e.g. Honeybees? Be great to cite an example!
% Note by Yan - if we actually have the sequence, then this turns into
% a sample node, so would be normally included anyway. Perhaps, therefore
% we should talk about these ``pass though'' nodes in the context
% of internal samples.
% Gregor: There are no good deeply sequenced pedigrees in honeybees, but sex
% chromosome behaves like haplodiploid system, also any deep pedigree works here, so I cited our pig
% work. Also, I think its important to not think of all genomes as current samples - we have many
% generations of animals with genome or SNP array data and we know pedigree so we can't assume they
% are all current samples - I believe you made this clear distinction also in the tsdate paper with
% ancient human samples

@@

\section*{Ancestry resolution}\label{Ancestry_resolution}
% Need to introduce Hudson's algorithm as concerned with simulating the process of
% the coalescent with recombination as efficiently as possible, and properly merge in
% with this section.

The edge annotations in Figs~\ref{fig-arg-data-structure} and \ref{fig-arg-in-pedigree}
describe the regions of genome that have been passed from a parent to its immediate children,
hence edges either span the entire genome, or all the region to the left (or right) of a breakpoint
(e.g. Fig~\ref{fig-arg-data-structure}C). In this form of edge annotation, which we call
``immediate edge annotation'',  % NB - perhaps we can think of a better term than this
it is inefficient to extract the local trees from the ARG structure.

There is, however, an alternative edge annotation scheme which allows efficient
storage and extraction of the local trees. It is based on the fact that the ARG
structure is inherently retrospective, defined as the ancestry of a set of \emph{sample}
nodes (such as \textsf{a}, \textsf{b}, and \textsf{c} in the examples above). We define a
``sample-resolved edge annotation'' to be a representation where edge spans
represent only those parts of the genome that have been inherited by the samples.
This also allows the removal of nodes which are not genetically ancestral to the samples.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
The \citet[][fig. 1]{wiuf1999recombination} example graph as a gARG, with contrasting
edge annotation schemes.
(A) Immediate edge annotation, reflecting ``local'' transmission of all
genetic material between parent and child. All edges span the full genome from position 0 to 7,
apart from those linking a recombinant and its two parents (bold).
(B) Sample-resolved edge annotation: the Hudson-like simplification algorithm has been used to
shorten edge spans to reflect only the upwards transmission of genetic
material from the samples \textsf{a}, \textsf{b}, and \textsf{c}. The depiction of the genome within
each node is shaded as in figure \ref{fig-arg-in-pedigree}a to show
coalescing regions of genome back to the grand MRCA genome \textsf{w}.
Line widths reflect the genomic span of each edge, which
helps visualise the relative importance of different transmission pathways.
}
\end{figure}

Fig~\ref{fig-ancestry-resolution} illustrates the two different edge annotation schemes.
To obtain the sample-resolved annotation in Fig~\ref{fig-ancestry-resolution}B we must
traverse backwards in time from sample nodes \textsf{a}, \textsf{b}, and \textsf{c}, using the
edge intervals to record regions of sample genome that have passed through each route.
When edges join into a parent node, the parent is allocated the union of the intervals.
When a child node is connected to two parents, one parent is allocated the edge
intervals of the child, truncated to include only genomic material to the left of the breakpoint.
Similarly the other parent is allocated the child's intervals truncated such that they contain
only regions to the right of the breakpoint (a natural extension exists for multiple
chromosomal breakpoints).

For example, in Fig~\ref{fig-ancestry-resolution}A the edge between child node \textsf{g} and
parent node \textsf{r} spans the entire genome from 0..7, reflecting the material
transmitted immediately between parent and child: i.e. local to those two nodes in the tree.
However, in Fig~\ref{fig-ancestry-resolution}B the edge has been reduced in extent so that
it now spans 0..2, the region that eventually ends up in
sample node \textsf{c}. This sample-resolved edge annotation scheme cannot be
derived locally: it must be done by traversing backwards in time from the sample nodes
up through the ARG.

% TECHNICAL DESCRIPTION
% Given a gARG $E$ and the set of samples $S$ we can \emph{resolve} its
% ancestry by noting that the resolved interval $Ir$ on an edge $(c,p,I)$
% is given by [Rough notes]
% \begin{itemize}
% \item The ``ancestry interval'' $A_u$ of a node $u$ is the union of
% all $R_{u, v}$ for all the children $v$ of $u$. (this isn't quite right)
% \item The resolved interval is then the intersection of $A_u$ and the
% original inheritance interval $I_v$ on the edge.
% \item It would be good of if differentiate between the inheritance intervals
% from the gARG and the ancestry intervals of the resolved gARG.
% \end{itemize}
% This process is illustrated in Fig~\ref{fig-ancestry-resolution}. It is intimately linked to
%Hudson's algorithm which operates by
% tracking segments of ancestral material carried by ancestors as
% we go backwards in time.

% jk-note unclear here about the best choice of notation. We could have
% lists of tuples (l, r, a), or a tuple of vectors (l_j, r_j, a_j),
% of a tuple (I_j, a_j), where I is a vector of intervals. Going to
% stick with the original tuple notation for now.
%
% Each lineage consists of a set of
% equal-length vectors $\mathbf{l}$, $\mathbf{r}$ and $\mathbf{a}$
% such that for each $1 \leq j \leq \left| l \right|$,
% $[\mathbf{l}_j, \mathbf{r}_j)$ is a half-closed genomic interval and
% $\mathbf{a}_j$ is an integer
% tracking the number of samples the lineage is ancestral to over the interval.
% Assume that the intervals are sorted from left-to-right so that $\mathbf{l}_1$
% and $\mathbf{r}_{|\mathbf{r}|}$ are the left- and right-most positions
% covered by the lineage, respectively.

When traversing upwards through the ARG like this, it is also helpful to record
the number of sample nodes to which a given section of genome is ancestral.
This is indicated in Fig~\ref{fig-ancestry-resolution}B by successively darker shading
of the regions within each node. When a region has all samples as descendants, it is
shaded black, and we can say that that region of the genome has fully coalesced.

Formally, we can define a sample-resolved gARG as a tuple $(S, E)$
where $S \subset \mathbb{N}$ is a set of sample nodes and
$E$ is a set of edges. Each element of $E$
is a tuple $(c, p, I)$ such that $c, p \in \mathbb{N}$,
$c$ is the child node, $p$ is the parent node and $I$ is the set of
disjoint genomic intervals $(\ell, r]$
over which genome $c$ inherits from $p$, and there is a path from
$c$ to some non-empty subset of $S$ through the local trees.

The ARG traversal algorithm which implements this edge trimming and removal process
is a form of ``simplification'' (cite), and can be thought of as running Hudson's algorithm
on a fixed tree topology (see appendix XXX).

% NB: below is a discussion of the Hudson algorithm which we might want to 
% retain in an appendix
% Each lineage consists of a list of
% disjoint ancestry segments $(\ell, r, a)$, where
% $[\ell, r)$ is a half-closed genomic interval and $a$ is an integer
% tracking the number of samples to which the lineage is ancestral over that interval.
% (We also usually track the tree node associated with each segment, but
% that is not important for our purposes here so we omit it.)
% If we have $n$ samples and a genome of length $m$, the process begins with $n$ lineages
% of the form $\{(0, m, 1)\}$. The process then works backwards in time from
% the present day as a series of random common ancestor or recombination events.
% Recombination events occur at a rate determined by the amount of ancestral material and
% the way in which it is distributed along a chromosome.
% Let $L$ be the set of ancestors at a given time $t$. Recombination events
% happen at rate $\rho \nu / (m - 1)$ where
% \[
% \nu = \sum_{x \in L}\left( \max_{(\ell, r, a) \in x}r
%     - \min_{(\ell, r, a) \in x}\ell - 1 \right)
% \]
% is the number of available `links' that may be broken. Thus, the rate of
% recombination is determined by the left- and right-most extent of the
% ancestral material carried by each lineage. At a recombination
% event we choose one of these links uniformly and break it. Given a lineage
% $x = [(\ell_j, r_j, a_j)]$ and a breakpoint $k$, we have two lineages
% $x_1$ and $x_2$ such that FILL IN DETAILS

% When $k = |L|$ lineages are present, common ancestor events
% occur at rate $\binom{k}{2}$. In a common ancestor event, two lineages
% are chosen uniformly at random and their ancestry segments merged.
% If we have overlapping intervals of ancestry from the two lineages,
% say, $(\ell, r, a_1)$ and $(\ell, r, a_2)$, a
% \emph{coalescence} occurs and ancestor represented by the current event
% will be present as a node (at least) in the marginal trees covering
% the interval $[\ell, r)$. The result of this coalescence is a segment
% $(\ell, r, a_1 + a_2)$, and if $a_1 + a_2 < n$ it is included in the
% ancestry for the new lineage. Otherwise, if $a_1 + a_2 = n$ we know that
% we have found the most recent common ancestor of all samples in
% the interval $[\ell, r)$ and so we do not need to simulate its history any further.
% Nonoverlapping intervals of ancestry from the two lineages are included
% in the resulting lineage without changes. Eventually, as the process continues,
% we find resultant lineages in which all segments have fully coalescenced,
% and so the number of extant lineages gradually dwindles down to zero.

Several features of Fig 3B are worthy of note. First, since node \textsf{j} is not on
the ancestral path from the samples it (and its associated edges) can be removed.
For larger ARGs, the absence of these nodes makes sample-resolved edge annotation
substantially more efficient than immediate edge annotation. Second, by the time we reach
nodes \textsf{t} and \textsf{v}, all regions of the genome have fully coalesced, and so
\textsf{w}, the ``grand MRCA'' can be omitted from the graph, at the cost of having
multiple root nodes in different regions of the genome. Third, there exist nodes such as
\textsf{k} with two children (i.e. ``common ancestor'' nodes in Hudson's terminology)
but in which no ancestral coalescence occurs. Fourth, some breakpoint positions occur
within regions of genome that are not ancestral (e.g. the breakpoint at position 3 in
node \textsf{k}). in this case sample-resolved edge annotations lose the exact position of
the breakpoint, although they retain the information that a breakpoint occurred between
positions 2 and 4 (which is sufficient for calculating the likelihood, see appendix XXX).

% Should probably define how we build the local trees somewhere.
TODO: explain why this encoding allows efficient extraction of local trees, as implemented in the tskit library.

@@

TODO: give some feel for how much of a reduction the sample-resolved encoding can give in a large ARG, and
maybe mention ``big'' vs ``little'' args as may be described in Jere's appendix

@@

% The state-space of the Griffiths process is much simpler than Hudson's algorithm,
% which greatly facilitates mathematical reasoning. This simplicity comes at a
% substantial cost, however, if we wish to use it as a practical means of
% simulating recombinant ancestry. The big ARG is vast, and any realisation
% for even moderate levels of recombination is far too large to be of practical
% use. The number of events in the big ARG all the way back to the GMRCA
% is $O(e^\rho)$~\citep{griffiths1997ancestral}, whereas the number
% of events required to simulate the little ARG is
% $O(\rho^2)$~\citep{hein2004gene,baumdicker2021efficient}.
% This disparity in the number of events in the two formulations is
% because the majority of the events that occur in the big ARG do
% not affect the genetic ancestry of the sample in any way. Recombination
% events that occur outside of ancestral material do not have any bearing
% on the ancestry of the sample, and so the structure is hugely redundant.
% As~\cite{wiuf1999recombination} note,
% ``an `ancestral' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.''
% Full quote:
% This process simplifies mathematics on the account that the notion of an
% ancestor will have a less restrictive meaning than usual:
% An ``ancestral'' sequence in the birth and death process
% need not have any genetic material in common with a
% sequence descended from it.

@@

Under the coalescent with recombination (or SMC) each local
tree along the genome is transformed into the next by
a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the ``subtree transfer'' operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.
